---
title: "Statistical analysis Agithu-3 data"
output: word_document
---

# Statistical analysis Agithu-3 data

The data used here has been previously filtered using Rfiltering_1.Rmd and manually curated.
Output files of Rfiltering_1.Rmd were:
* 0.975_count10_cleanrepeats_motus.csv
* 0.975_count10_cleanrepeats_reads.csv
* 0.975_count10_cleanrepeats_samples.csv

Manual curation was performed to mitigate potential biases. A taxonomic expert checked the taxonomic annotations to assess the likeliness of occurrence in the region at the time. Based on this assessment, several taxonomic assignments were adjusted and we added a column "to_exclude" with TRUE or FALSE values to the motu table. This was set to TRUE for sequences with <98% ID match, without family annotation, or deemed unlikely (see SOM table S3).
We also added further metadata on the samples to the samples table and renamed the adjusted .csv files to:
* DNAplants.csv
* DNAdata.csv
* DNAsamples.csv


## Prepare the R environment

```{r}
library(tidyverse)
library(vegan)
library(ggpubr)
library(wesanderson)
```

## Loading of the dna data

```{r}
dna_plants = read.csv2(file="../data/DNAplants.csv", row.names = 1, na.strings=c(""," ","NA"))
dna_repsdata = read.csv2(file="../data/DNAdata.csv", row.names = 1)
dna_samples = read.csv2(file="../data/DNAsamples.csv", row.names = 1)

#ensure the order of the plants are the same in both tables
dna_repsdata = dna_repsdata[order(match(rownames(dna_repsdata), rownames(dna_plants))), ]

# cast as numeric instead of character
dna_repsdata = as.data.frame(lapply(dna_repsdata, as.numeric))
rownames(dna_repsdata) = rownames(dna_plants) # re-add the rownames

#ensure the order of the samples are the same in both tables
dna_repsdata = dna_repsdata[order(match(rownames(dna_repsdata), rownames(dna_samples))), ]

dim(dna_plants)
dim(dna_repsdata) # plants as columns, replicates as rows
dim(dna_samples)
rownames(dna_samples) = dna_samples$sample_id
```

```{r}
# see how many replicates we have per sample
# let's extract the sample names from the reads table
names = as.character(colnames(dna_repsdata))
samplenames_split = strsplit(names,"[_]")

# save the first part as the sample name
samplenames = sapply(samplenames_split,
                   function(x) x[1])

replicates = sapply(samplenames_split,
                   function(x) x[length(x)])

dna_reps = data.frame(sample = samplenames,
                      replicate = replicates)
head(dna_reps)
table(dna_reps$sample)
```
## Further filtering steps


### Remove plants present in < 2 PCR replicates
```{r}
# prepare a binary table
dna_repsbin = dna_repsdata
dna_repsbin[dna_repsbin > 0] <- 1

# remove plants only present in 1 rep
dna_plants = dna_plants[rowSums(dna_repsbin)>1, ]
dna_repsdata = dna_repsdata[rowSums(dna_repsbin)>1, ] 
dim(dna_repsdata)
sum(rowSums(dna_repsdata))
```

```{r}
# remove plants we want to exclude, as per 'to_exclude' column:
# with <98% ID match,
# without family annotation, 
# or deemed unlikely by a taxonomic expert

dna_repsdata = dna_repsdata[dna_plants$to_exclude == FALSE, ]
dna_plants = dna_plants[dna_plants$to_exclude == FALSE, ] 

# remove replicates that are now empty
dna_repsdata = dna_repsdata[,colSums(dna_repsdata)!=0] 

dim(dna_repsdata)
sum(rowSums(dna_repsdata))
```

```{r}
write.csv(dna_repsdata, "../results/dna_dataperreplicate_cleaned.csv")
write.csv(dna_plants, "../results/dna_plants_cleaned.csv")
```

### getting the sample information to merge the filtered replicates
```{r}
# let's extract the sample names from the reads table
names = as.character(colnames(dna_repsdata))
samplenames_split = strsplit(names,"[_]")

# save the first part as the sample name
samplenames = sapply(samplenames_split,
                   function(x) x[1])

replicates = sapply(samplenames_split,
                   function(x) x[length(x)])

dna_reps = data.frame(sample = samplenames,
                      replicate = replicates)
```

```{r}
# make a temporary dataframe to save the number of replicates per sample
rs = as.data.frame(table(dna_reps$sample))

# check which samples are missing
missingrows = setdiff(rownames(dna_samples),rs$Var1)
missingrows = data.frame( "Var1" = missingrows,
                          "Freq" = rep(0, length(missingrows))
                          )

# add the missing samples
rs = rbind(rs, missingrows)
rownames(rs) = rs$Var1

# reorder the table to match the rownames
rs_ordered = rs[order(match(rs$Var1, rownames(dna_samples))), ]

# add the number of replicates per sample to the samples dataframe
dna_samples$replicates = rs_ordered$Freq
```


### Merge the replicates
```{r}
# prepare a binary table
dna_repsbin = dna_repsdata
dna_repsbin[dna_repsbin > 0] <- 1

# merge the DNA data sample replicates to get a reads per sample table
dna_data = aggregate(t(dna_repsdata), list(dna_reps$sample), FUN=sum)
rownames(dna_data) = dna_data$Group.1
dna_data$Group.1 = NULL

# merge the binary DNA data sample replicates to get a replicates per sample table
dna_repdata = aggregate(t(dna_repsbin), list(dna_reps$sample), FUN=sum)
rownames(dna_repdata) = dna_repdata$Group.1
dna_repdata$Group.1 = NULL
```

```{r}
# set samples with only one replicate to 0
#to_remove = rownames(dna_samples[dna_samples$replicates == 1,])
to_remove = which(rowSums(dna_repdata) == 1)
dna_data = dna_data[!rownames(dna_data) %in% to_remove,]
dna_repdata = dna_repdata[!rownames(dna_repdata) %in% to_remove,]
dim(dna_data)
sum(rowSums(dna_data))
```
```{r}
#  remove plants only in 1 rep
dna_data = dna_data[ , colSums(dna_repdata)>1 ] 
dna_plants = dna_plants[colSums(dna_repdata)>1, ]
dna_repdata = dna_repdata[ , colSums(dna_repdata)>1] 

dim(dna_plants)
dim(dna_data)
dim(dna_repdata)
dim(dna_samples)
sum(rowSums(dna_data))
```
```{r}
# prepare a missing sample row with only 0s
empty=matrix(c(rep.int(0,length(dna_data))),nrow=length(setdiff(dna_samples$sample_id, rownames(dna_data))),ncol=length(dna_data))  
colnames(empty) = colnames(dna_data)  
rownames(empty) = setdiff(dna_samples$sample_id, rownames(dna_data))

# add the missing sample to the DNA data tables
dna_data = rbind(dna_data, empty)
dna_repdata = rbind(dna_repdata, empty)

# make sure the names match for the tables
dna_data = dna_data[match(rownames(dna_samples), rownames(dna_data)), match(rownames(dna_plants), colnames(dna_data))]
dna_repdata = dna_repdata[match(rownames(dna_samples), rownames(dna_repdata)), match(rownames(dna_plants), colnames(dna_repdata))]

# check the total number of reads
sum(rowSums(dna_data))

write.csv(dna_data, "../results/dna_datareads_clean.csv")
write.csv(dna_repdata, "../results/dna_datareps_clean.csv")
write.csv(dna_plants, "../results/dna_plants_clean.csv")
```