---
title: "Filtering trnl tabular file"
output: word_document
---

# Aghitu-3 data filtering

This Rmarkdown file uses the results from OBITools processing as input and further filters the metabarcoding data.
The code provided here is a mash-up from different sources, including the ForBio course: DNA Metabarcoding: Data processing and interpretation.
Special thanks and credit goes to the teachers: 
Eric Coissac
Frederic Boyer
Youri Lammers




# Preparation before filtering 

```{r}
# Prepare the R environment
require(tidyverse)
```

## Setting the filters
```{r}
# Filter thresholds to be adjusted according to experiment
bestid_thr = 0.975 # for best identity match with reference database
thrCount_motu <- 10 # minimal number of reads to keep a MOTU
```

## Loading the data
```{r}
raw = read.delim(file = "../data/assigned_taxo_clean005_ann.tab")
```

In 'raw' the OTUs are in rows and all of the information (quality, taxonomy, readcounts, etc.) in columns
Here we prepare a transformed version of the OTU table and separate it into specific tables:
 - samples
 - motus
 - reads
 
```{r}
# motus should be a dataframe with the metadata for the motu ids
# so we need a subset of the 'raw' dataframe without the readcounts
motus = select(raw, -(starts_with('sample')))
# and the opposite subset for the reads table
reads = select(raw, starts_with('sample'))
# change NAs (reads found for one database but not the other) into zeros 
# NOTE: this is not necessary for the mammal data
reads[is.na(reads)] = 0
rownames(motus) = motus$id
rownames(reads) = motus$id

# Samples should just be a dataframe with complete samplenames as rownames and as a column "Samples"
# so let's extract the sample names from the reads table
names = as.character(colnames(reads))
samplenames_split = strsplit(names,"[.]")

# save the last part as the sample name
samplenames = sapply(samplenames_split,
                   function(x) paste0("AGH", x[length(x)]))
samples = data.frame(sample = samplenames)
rownames(samples) = samplenames

# make sure the names match for the samples and reads tables
colnames(reads) = samplenames 

reads = as.data.frame(t(reads)) # make the rows = samples and columns = MOTUs
```

```{r}
# check the dimensions of the resulting dataframes
dim(motus)
dim(samples)
dim(reads)
sum(rowSums(reads))
```


### extract sample names
```{r}
# get sample names
sample_names = as.character(samples$sample)

# Split the names on the underscores
sample_names_split = strsplit(sample_names,"_")
#head(sample_names_split)

# save the last part as replicate
replicate = sapply(sample_names_split,
                   function(x) x[length(x)])

# save the first part as the sample id
sample_id = sapply(sample_names_split,
                   function(x) x[1])

sample_type = sapply(sample_names_split,
                   function(x) x[2])
# label everyting as sample, except the controls
sample_type[grep(pattern = "^s",sample_type)]="sample"
sample_type[grep(pattern = "^pnc",sample_type)]="pcrneg"
sample_type[grep(pattern = "^enc",sample_type)]="extneg"
sample_type[is.na(sample_type)]="leakage" # we found some unused tags such as "trnL170"
```

```{r}
# builds a data frame with the sample descriptions
sample_desc = data.frame(name      = sample_names,
                         sample_id = sample_id,
                         replicate = replicate,
                         type      = sample_type,
                         nonReplicating = FALSE # needed later on for filtering
                         )
rownames(sample_desc) = sample_desc$name

# Substitute the old sample description by the new one
samples = sample_desc
```


### Add MOTU information
```{r}
sequence_type = rep("Unknown",nrow(motus))

# check the best identity of the sequences and label them as embl when above the threshold
sequence_type[motus$`best_identity.trnl_r142` >= bestid_thr] = "embl" 

motus$sequence_type=factor(sequence_type)
motus$counts_in_blanks = sapply(reads[samples$type != "sample",], sum)
motus$in_blanks = motus$counts_in_blanks > 0
```


## Having a look at the data 

```{r}
df = motus[, c("best_identity.trnl_r142", "species_list.trnl_r142")]
df$species_list = strsplit(df$species_list.trnl_r142, ", ")[]
df$species_number = as.numeric(unlist(lapply(df$species_list, length)))
df$best_id = as.numeric(df$best_identity.trnl_r142)

require(ggplot2)
require(ggpubr)
ggplot(df, aes(x=best_id, y=species_number)) + 
  geom_point() +
  geom_smooth(method="lm")+
  theme_classic2() +
  theme(axis.text = element_text(colour = "black"))

# or alternatively:
require(ggstatsplot)
ggscatterstats(df, best_id, species_number, type="bayesian")
```

### Basic descriptions of the samples
```{r}
#The dataset is constituted of 
nrow(reads) #PCRs replicates for
ncol(reads) #MOTUs 

# Counting the reads per PCR reaction
reads_per_pcr = rowSums(reads)

# Counting the MOTUs per PCR reaction
motus_per_pcr = rowSums(reads > 0)

# Counting the reads per sample
reads_per_sample = aggregate(reads_per_pcr, list(sample_id), FUN=sum)
reads_per_sample[order(reads_per_sample$x, decreasing = FALSE),]

# Counting the MOTUs per sample
motus_per_sample = aggregate(motus_per_pcr, list(sample_id), FUN=sum)
motus_per_sample[order(motus_per_sample$x, decreasing = FALSE),]
```


```{r}
# taking a look at the distribution
par(mfrow=c(1,2))
hist(log10(reads_per_pcr),
     breaks=50,
     main = "Reads per PCR",
     xlab="Reads count (log scale)")
hist(motus_per_pcr,breaks=50,
     main = "MOTUs per PCR",
     xlab="MOTU count")
```


## cleaning up the annotations we no longer need
```{r}
# annotations we want to keep
idx= c("id", "count", "best_identity.trnl_r142",
       "family_name", "genus_name", "species_name", 
       "scientific_name", "seq_length", "sequence", "species_list.trnl_r142", "counts_in_blanks","in_blanks")
motus = motus[, idx]

# rename 
motus$best_id = motus$`best_identity.trnl_r142`
motus$`best_identity.trnl_r142` = NULL

# most abundant MOTUs on top
idx = order(colSums(reads), decreasing=T)
motus = motus[idx,]
reads = reads[,idx]
colnames(reads) = motus$id

samples = samples[match(rownames(reads), rownames(samples)),]
```


## Identify artifacts

### Reagent contaminants

Identification of OTUs of maximum abundance in the extraction and pcr blanks

```{r, results="hide", warning=FALSE}
stopifnot(all(rownames(samples)==rownames(reads)))

maxInExtractionCtrl <- apply(reads[samples$type=='extneg',], MARGIN=2, function(x) max(x,na.rm = T))
maxInPCRCtrl        <- apply(reads[samples$type=='pcrneg',], MARGIN=2, function(x) max(x,na.rm = T)) 
maxInSamples        <- apply(reads[samples$type=='sample',], MARGIN=2, function(x) max(x,na.rm = T)) 

df <- data.frame(maxInExtractionCtrl, maxInPCRCtrl, maxInSamples)

#determining the type of problematic OTU
motus$bias <- c('extneg','pcrneg',NA)[apply(df, MARGIN=1, FUN=which.max)]

#adding OTUs infos
infosCols <- c("count", "best_id", "family_name", "genus_name","species_name", "bias")

df <- cbind(df, motus[,infosCols])

#keeping only problematic sequences for display

df <- df[!is.na(df$bias),]
df <- df[order(df$maxInExtractionCtrl+df$maxInPCRCtrl, decreasing=T),]
```


### Visualizing the level of contaminants in the samples

```{r}
stopifnot(all(colnames(reads)==motus$id))

if (dim(df)[1] > 1) {
  plot(rowSums(reads[, !is.na(motus$bias)]), rowSums(reads[,!is.na(motus$bias)]>0), 
     col=as.factor(samples[rownames(reads), 'type']),
     log='xy', pch=16, xlab='#Reads', ylab='#OTUs')

  legend('topleft', 
       legend = levels(as.factor(samples[rownames(reads), 'type'])),
       col=1:nlevels(as.factor(samples[rownames(reads), 'type'])), 
       pch=16, cex=0.5)
} else {
  plot(sum(reads[, !is.na(motus$bias)]), sum(reads[,!is.na(motus$bias)]>0), 
     col=as.factor(samples[rownames(reads), 'type']),
     log='xy', pch=16, xlab='#Reads', ylab='#OTUs')

  legend('topleft', 
       legend = levels(as.factor(samples[rownames(reads), 'type'])),
       col=1:nlevels(as.factor(samples[rownames(reads), 'type'])), 
       pch=16, cex=0.5)
}

```


# Filtering steps

## Removing the contaminants

```{r}
table(motus$bias)
```

```{r}
correctedCounts2 <- reads

correctedCounts2[,motus$in_blanks] <- 0  ## removing contaminants based merely on presence in blanks

# check how many are removed; FALSE means they're empty; TRUE means that they're kept
table(colSums(correctedCounts2) > 0)


plot(rowSums(reads), rowSums(correctedCounts2), xlab='#Reads before cleaning for contaminants',
     ylab='#Reads after cleaning for contaminants', main='Effect of contaminants removal\non #Reads',
     col=as.factor(samples[rownames(reads), 'type']), pch=16)
abline(a=0,b=1)

legend('topleft', 
       legend = levels(as.factor(samples[rownames(reads), 'type'])),
       col=1:nlevels(as.factor(samples[rownames(reads), 'type'])), 
       pch=16, cex=0.5)

plot(rowSums(reads>0), rowSums(correctedCounts2>0), xlab='#OTUs before cleaning for contaminants',
     ylab='#OTUs after cleaning for contaminants', main='Effect of contaminants removal\non #OTUs',
     col=as.factor(samples[rownames(reads), 'type']), pch=16)
abline(a=0,b=1)

legend('topleft', 
       legend = levels(as.factor(samples[rownames(reads), 'type'])),
       col=1:nlevels(as.factor(samples[rownames(reads), 'type'])), 
       pch=16, cex=0.5)

```

```{r}
sum(colSums(correctedCounts2) > 0) # total number of motus left
sum(rowSums(correctedCounts2)) # total number of reads left
```
```{r}
# updating the tables
correctedCounts2 = correctedCounts2[rowSums(correctedCounts2) > 0, colSums(correctedCounts2) > 0] # filter out empty sequences and samples
motus = motus[colnames(correctedCounts2),]
samples = samples[rownames(correctedCounts2),]
dim(correctedCounts2)
```


## Removing bad MOTUs

Bad MOTUs are those: 
- that score below a specified identity threshold
- occurring with < 10 reads in the entire dataset


### ID threshold < XX % match
```{r}
# keeping track of things
table(colSums(correctedCounts2) > 0) # TRUE are those that are kept

# determining which are below the bestid_thr and removing those from the reads
motus$below_bestid_thr = motus$best_id<bestid_thr
correctedCounts2[,motus$below_bestid_thr] <- 0 

# now check again, the number for TRUE should be smaller
table(colSums(correctedCounts2) > 0)

sum(colSums(correctedCounts2) > 0) # total number of motus left
sum(rowSums(correctedCounts2)) # total number of reads left
```
```{r}
# updating the tables
correctedCounts2 = correctedCounts2[rowSums(correctedCounts2) > 0,colSums(correctedCounts2) > 0] # filter out empty sequences and samples
motus = motus[colnames(correctedCounts2),]
samples = samples[rownames(correctedCounts2),]

dim(correctedCounts2)
```


### Sequence/MOTU minimal count filter threshold < 10 reads in a PCR repeat
```{r}
# determining which cells are below a certain read count and removing those from the reads
correctedCounts2[correctedCounts2<thrCount_motu] = 0 # set cell to 0

# check how many
table(colSums(correctedCounts2) > 0)
```
```{r}
# save and check what is left
reads1 = correctedCounts2[rowSums(correctedCounts2) > 0, colSums(correctedCounts2) > 0] # filter out empty sequences and samples
motus1 = motus[colnames(reads1),]
samples1 = samples[rownames(reads1),]

# check number of replicates, motus and reads
dim(reads1)
sum(colSums(reads1)) # total number of reads left
```

## Removing PCR repeats with < .... reads in total
```{r}
thrCount_pcr = 10
# removing those with low amount of reads / empty ones
l <- lapply(levels(as.factor(samples[rownames(reads1), 'type'])), function(st) {
  if (is.null(dim(reads1[samples[rownames(reads1), 'type']==st,]))) {
    sum(reads1[samples[rownames(reads1), 'type']==st,])
  } else {
  rowSums(reads1[samples[rownames(reads1), 'type']==st,])
  }
})
names(l) <- levels(as.factor(samples[rownames(reads1), 'type']))

boxplot(l, las=2, cex=0.25, outpch=NA, main='#reads before filtering')
stripchart(l, vertical = T, 
           method="jitter", pch=16, cex=0.4, add=T)
abline(h=thrCount_pcr, col='red')
```

```{r}
## labelling samples that are below a threshold readcount
samples1$empty_PCR <- FALSE

stopifnot(all(rownames(samples1)==rownames(reads1)))
samples1$empty_PCR[rowSums(reads1)<thrCount_pcr] <- TRUE

## filtering out those that are "empty" (both PCR repeats as sequences)
correctedCounts3 = reads1[!samples1$empty_PCR, colSums(reads1) > 0]

reads2 = correctedCounts3[rowSums(correctedCounts3) > 0,] ## not sure this step is needed, but doesn't hurt
motus2 = motus[colnames(reads2),]
samples2 = samples[rownames(reads2),]
```

```{r}
# how many replicates, motus and reads do we have left?
dim(reads2)
```
```{r}
sum(colSums(reads2) > 0) # total number of motus left
sum(rowSums(reads2)) # total number of reads left
```

# Save the resulting files

```{r}
write.csv(motus2,file = paste0("../results/", bestid_thr, "_count", thrCount_motu,"_cleanrepeats_motus.csv"))
write.csv(reads2,file = paste0("../results/", bestid_thr, "_count", thrCount_motu,"_cleanrepeats_reads.csv")) 
write.csv(samples2,file = paste0("../results/", bestid_thr, "_count", thrCount_motu,"_cleanrepeats_samples.csv")) 
```


## getting some basic understanding of what is left

```{r}
# which samples are still there?
reads_per_sample = rowSums(reads2)
motus_per_sample = rowSums(reads2 > 0)

plot(reads_per_sample,motus_per_sample,
     xlab="Read counts",
     ylab="MOTU counts",
     col=as.factor(samples2$type),
     pch=19, cex=1,
     log="xy"
     )

# create a transparent color
mygrey <- rgb(20, 20, 20, max = 255, alpha = 100, names = "black") # alpha is 0-255

text(reads_per_sample,motus_per_sample, labels=rownames(reads2), col = mygrey, cex=0.6, font=2)

```

```{r}
# how many samples have more than 1 motu?
length(motus_per_sample[motus_per_sample>1]) 

# how many samples have more than 5 motus?
length(motus_per_sample[motus_per_sample>5])
```
```{r}
table(samples2$sample_id)
```


```{r}
# get summary information about the plants that we found
tmp = unique(motus2$family_name)
motus2$family_name = as.character(motus2$family_name)
motus2$family_name[is.na(motus2$family_name)] <- "unknown"

# calculate the number of reads per motu
motus2$count = colSums(reads2)

# calculate the number of reads per family
reads_per_fam = aggregate(motus2$count, list(family = motus2$family_name), sum)
reads_per_fam[order(reads_per_fam$x, decreasing = TRUE),]

# calculate the number of reads per genus
reads_per_gen = aggregate(motus2$count, list(genus = motus2$genus_name), sum)
reads_per_gen[order(reads_per_gen$x, decreasing = TRUE),]

# calculate the number of reads per species
reads_per_spec = aggregate(motus2$count, list(species = motus2$species_name), sum)
reads_per_spec[order(reads_per_spec$x, decreasing = TRUE),]
```
